---
title: "PACE breast cancer analysis (Figure 2)"
author: "Elijah Willie"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

This notebook reproduces the breast cancer analysis and Figure 2 of the PACE
manuscript **using the [PACE package](https://github.com/ecool50/PACE)**. The
method (neighbour kernels, per-cell contamination correction, NB1 PQL fit, mashr
shrinkage, variance decomposition, driver scores) and the standard figures come
from the package; only the bespoke spatial ROI-inset composition is a local
helper. Rendered from the repository root:

```bash
quarto render pace_mv/notebooks/analysis_bc.qmd
```

# Setup

```{r setup}
library(PACE)
library(SpatialExperiment)
library(SummarizedExperiment)
library(Matrix)
library(dplyr)
library(ggplot2)

# Bespoke ROI-inset figure (Figure 2e) -- not part of the PACE package.
source("../helpers/compose_inset_figure.R")

spe_path <- "../../data/spe_10x_nuclei_withMetrics.rds"   # fetched by fetch_data.R
```

# Load and prepare the section

Collapse the 17 scClassify sub-states into the eight modelled cell types, drop
the rare mast population, and keep genes detected in at least 5% of cells of some
cell type -- exactly the manuscript filter (126,192 cells x 275 genes).

```{r prepare}
spe <- readRDS(spe_path)

celltype_map <- c(
  `CRABP2+ Malignant`="Tumour", `ECM1+ Malignant`="Tumour", `SCGB2A2+ Malignant`="Tumour",
  `ACTA2+ Myoepi`="Myoepithelial", `KRT15+ Myoepi`="Myoepithelial",
  `Firoblast`="Stromal", `VWF+ Endothelial`="Endothelial", `STAB2+ Endothelial`="Endothelial",
  `CD4 T`="T_Cell", `CD8 T`="T_Cell", `B Cells`="B_Cell", `Plasma`="B_Cell",
  `Macrophage`="Macrophage", `CD163+ Macrophage`="Macrophage",
  `IRF7+ DC`="Dendritic_Cell", `LAMP3+ DC`="Dendritic_Cell", `Mast cells`="Mast")
# canonical cell-type order: indexes the random-effect blocks of the locked fit
cell_types <- c("T_Cell", "Stromal", "Macrophage", "Tumour", "Endothelial",
                "B_Cell", "Myoepithelial", "Dendritic_Cell")

ct   <- celltype_map[as.character(colData(spe)$scClassify)]
keep <- !is.na(ct) & ct != "Mast"
spe  <- spe[, keep]
spe$cellType <- factor(ct[keep], levels = cell_types)

# drop zero-count cells, then per-type detection filter (>= 5% in some type)
spe <- spe[, colSums(assay(spe, "counts")) > 0]
counts <- assay(spe, "counts"); ctv <- as.character(spe$cellType)
det <- vapply(cell_types, function(t)
  as.numeric(rowMeans(counts[, ctv == t, drop = FALSE] > 0)), numeric(nrow(spe)))
spe <- spe[apply(det, 1, max) >= 0.05, ]

sprintf("%d cells x %d genes across %d cell types", ncol(spe), nrow(spe),
        nlevels(spe$cellType))
```

# Fit PACE

A single `paceFit()` call runs the whole pipeline. On the full section this takes
a few minutes on four cores. `types` fixes the cell-type block order and
`threads = 4L` the reduction order of the threaded solve; both are needed to
reproduce the locked fit rather than merely an equivalent one.

```{r fit, cache = TRUE}
fit <- paceFit(spe,
               celltype_col      = "cellType",
               image_col         = "slide",
               types             = cell_types,
               contamination     = "percell_hc",   # per-cell contamination correction
               dispersion        = "nb1",
               h_bio = 30, h_tech = 5,
               drop_sparse_neff  = 30,
               within_image      = TRUE,
               edge_correct      = TRUE,
               data_informed_tau = TRUE,
               n_iter = 32L, threads = 4L, tau_shrinkage = "adaptive",
               verbose = FALSE)
fit
```

The macrophage–tumour flagships reproduce the manuscript (MRC1 −0.156, APOC1 +0.121):

```{r flagships}
neighbourSlopes(fit) |>
  filter(focal == "Macrophage", neighbour == "Tumour",
         gene %in% c("MRC1", "APOC1")) |>
  select(gene, estimate_shrunk, lfsr)
```

# Figure 2a — per-focal variance decomposition

```{r fig2a, fig.width = 11, fig.height = 4.5}
plotDecomposition(fit)
```

# Figure 2b — pairwise spatial interactions

```{r fig2b, fig.width = 6.5, fig.height = 5}
plotPairHeatmap(fit)
```

# Figure 2c — macrophage–tumour drivers

```{r fig2c, fig.width = 10, fig.height = 4.5}
plotDrivers(fit, "Macrophage", "Tumour")
```

The other three top pairs from the heatmap:

```{r fig2c-other, fig.width = 10, fig.height = 4.5}
plotDrivers(fit, "Stromal", "Tumour")
plotDrivers(fit, "B_Cell", "T_Cell")
plotDrivers(fit, "Tumour", "Myoepithelial")
```

# Figure 2d — expression versus tumour proximity

```{r fig2d, fig.width = 11, fig.height = 4.6}
plotProximity(fit, spe, c("MRC1", "APOC1"), "Macrophage", "Tumour")
```

# Figure 2e — spatial ROI insets

Two tumour-margin regions of interest, macrophages coloured by MRC1/APOC1 over the
tumour background (the bespoke `compose_inset_figure()` helper).

```{r fig2e, fig.width = 12, fig.height = 7}
cp10k <- log1p(t(t(as.matrix(assay(spe, "counts"))) /
                 pmax(colSums(assay(spe, "counts")), 1) * 1e4))          # genes x cells
plot_df <- data.frame(
  x        = spatialCoords(spe)[, 1],
  y        = spatialCoords(spe)[, 2],
  celltype = as.character(spe$cellType),
  MRC1     = cp10k["MRC1", ],
  APOC1    = cp10k["APOC1", ])

compose_inset_figure(
  plot_df, genes = c("MRC1", "APOC1"), focal = "Macrophage", neighbour = "Tumour",
  xlim1 = c(6300, 7300), ylim1 = c(3700, 4700),
  xlim2 = c(6500, 7500), ylim2 = c(2300, 3300))
```

# Supplementary — contamination anchor genes

```{r anchors}
anchorGenes(fit, spe)
```

# Session info

```{r sessioninfo}
sessionInfo()
```
