---
title: "PACE-MV — Breast Cancer (10x Xenium) analysis"
author: "Elijah Willie"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    fig-width: 8
    fig-height: 5
    df-print: paged
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

Janesick 2023 breast cancer Xenium, analysed end-to-end with the
**current canonical PACE-MV pipeline** (per-cell-HC contamination model,
NB1 dispersion, uncapped frNN kernels). The starting point is a
`SpatialExperiment` (`spe_10x_nuclei_withMetrics.rds`) carrying
scClassify cell-type labels.

This is a **single physical section** (the `slide` column = FOV tiles /
CV folds, not biological replicates), so there is no disease contrast:
the inferential target is `V_spatial` — the spatial proximity slope
itself — not a condition × spatial interaction.

This notebook reproduces the locked fit
`data/breast_cancer/sweeps/mvpql_percell_hc.rds`. It refits from the SPE
rather than loading the cached bundle, so it is self-contained; the
numbers reproduce the canonical fit closely (the locked fit reads the
pre-built `Y_df_for_mcsd.rds` gene bundle, so tiny gene-set differences
are possible).

Modelled on the legacy `breast_cancer_pace_10x.Rmd` analysis arc,
modernised to the per-cell-HC model and the current `scripts/` tree.

# Setup

Load the package code (`scripts/zzz.R` auto-sources `scripts/core/` then
`scripts/helpers/`) and set the canonical model knobs. These are the
same constants the canonical builder uses; none are tuned per cohort
except `disp_model = "nb1"` (canonical) and the frNN kernel.

| Constant | Meaning |
|----|----|
| `h_tech = 5 µm` | exponential kernel scale (optical bleed) |
| `h_bio = 30 µm` | Gaussian kernel scale (biological proximity) |
| `kernel_radius = 3·h_bio = 90 µm` | frNN radius (uncapped neighbour sum) |
| `min_detection = 0.05` | gene QC: detected in ≥5% of ≥1 cell type |
| `min_neff = 30` | drop (focal,nb) kernel cols with effective n \< 30 |
| `n_iter = 32` | PQL outer iterations |
| `disp_model = "nb1"` | NB1 variance `μ(1+α)` (canonical) |

```{r setup}
rm(list = ls())
gc()

suppressPackageStartupMessages({
  # tidyverse + plotting
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(patchwork)
  # matrices, parallelism, neighbour search
  library(Matrix)
  library(BiocParallel)
  library(FNN)
  library(dbscan)
  # model + shrinkage
  library(glmmTMB)
  library(mashr)
  # data containers
  library(SpatialExperiment)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
})
options(dplyr.summarise.inform = FALSE)

# Run from pace_mv/ wherever this notebook is launched: here::here() finds the
# repo root (via lateral_spillover.Rproj), then we descend into pace_mv/.
setwd("../../pace_mv")

# scripts/zzz.R auto-sources scripts/core + scripts/helpers and compiles the C++;
# then the streaming solver and the pipeline.
source("scripts/zzz.R")                              # auto-loads scripts/ (core + helpers + C++)
source("scripts/core/06a-plots-decomp.R")            # decomp plot fns: zzz's numbered loader skips 06a/b/c/d
source("streaming/helpers/pace_mvpql_streaming.R")   # streaming solver (lives outside scripts/helpers/)
source("stage0/pace_core.R")                         # the pipeline functions
```

```{r}
# ---- canonical model knobs (BC per-cell-HC, NB1, frNN) ----
h_tech        <- 5            # optical-bleed kernel scale (µm)
h_bio         <- 30           # biological-proximity kernel scale (µm)
kernel_radius <- 3 * h_bio    # frNN truncation radius (µm); kernels < 1e-3 beyond
min_detection <- 0.05         # gene QC: detected in >= 5% of cells of >= 1 type
min_neff      <- 30           # drop (focal, neighbour) kernel cols with eff. n < 30
n_iter        <- 32           # PQL outer iterations
tau_shrinkage <- "adaptive"   # empirical-Bayes pooling of variance components
disp_model    <- "nb1"        # NB1 dispersion: Var = mu (1 + alpha) 
pql_threads   <- 4

# The 8 cell types PACE fits against (sub-states collapsed; see next section).
# Mast is EXCLUDED: only 240 cells (0.2%) -> too few to estimate spatial slopes,
# and as a rare neighbour it produced unstable, low-n flagship artifacts
# (e.g. Tumour<-Mast, n_eff ~43 vs ~300 for real pairs). Dropped at the QC filter.
cell_types <- c("T_Cell", "Stromal", "Macrophage", "Tumour", "Endothelial",
                "B_Cell", "Myoepithelial", "Dendritic_Cell")

# Inputs / outputs.
spe_path    <- "../../data/spe_10x_nuclei_withMetrics.rds"   # raw SPE (shared data lake)
output_path <- "../../data/breast_cancer/sweeps/mvpql_percell_hc_notebook.rds"
```


The model + reporting come from `stage0/pace_core.R` (`pace_fit_streaming()`,
`pace_shrink()`, `pace_decompose()`, `pace_top_drivers()`). The figure helpers
(`density_bin_boxplot()`, `top_gene()`, `fp_curve()`, `compose_inset_figure()`)
live in `scripts/helpers/` and are auto-sourced in the setup chunk.

# Load SPE and map to the 8-cell-type scheme

The 17 fine-grained scClassify labels carry sub-state detail PACE does
not need (three malignant states pool into `Tumour`, two myoepithelial
into `Myoepithelial`, etc.). Pooling keeps each focal random-slope block
well-powered. `unassigned` cells are dropped.

```{r load-spe}
spe <- readRDS(spe_path)
cat(sprintf("%d genes x %d cells\n", nrow(spe), ncol(spe)))

# Collapse the 17 scClassify sub-states; Mast is dropped at the filter below -> 8 modelled types.
celltype_map <- c(
  `CRABP2+ Malignant`  = "Tumour",
  `ECM1+ Malignant`    = "Tumour",
  `SCGB2A2+ Malignant` = "Tumour",
  `ACTA2+ Myoepi`      = "Myoepithelial",
  `KRT15+ Myoepi`      = "Myoepithelial",
  `Firoblast`          = "Stromal",
  `VWF+ Endothelial`   = "Endothelial",
  `STAB2+ Endothelial` = "Endothelial",
  `CD4 T`              = "T_Cell",
  `CD8 T`              = "T_Cell",
  `B Cells`            = "B_Cell",
  `Plasma`             = "B_Cell",
  `Macrophage`         = "Macrophage",
  `CD163+ Macrophage`  = "Macrophage",
  `IRF7+ DC`           = "Dendritic_Cell",
  `LAMP3+ DC`          = "Dendritic_Cell",
  `Mast cells`         = "Mast"
)

sc_label <- as.character(colData(spe)$scClassify)
celltype <- celltype_map[sc_label]
keep_cell <- !is.na(celltype) & celltype != "Mast"   # drop unassigned + Mast (240 cells; low-n)
spe       <- spe[, keep_cell]
celltype  <- celltype[keep_cell]
print(table(celltype))
```

# Build `Y` + `df`

`Y` = integer counts (cells × genes); `df` = cell metadata (`x`, `y`,
`celltype`, `slide`, `nCount`). Zero-count cells are dropped (they break
the `log(nCount)` offset).

```{r assemble}
# PACE convention: cells in rows, genes in columns.
Y <- t(as.matrix(assay(spe, "counts")))
storage.mode(Y) <- "integer"

cell_data <- colData(spe)
cell_xy   <- spatialCoords(spe)
df <- data.frame(
  cell_id  = as.character(cell_data$cell_id),
  x        = cell_xy[, 1],
  y        = cell_xy[, 2],
  celltype = factor(celltype, levels = cell_types),
  slide    = as.character(cell_data$slide),
  stringsAsFactors = FALSE
)

# Drop zero-count cells (log offset undefined).
df$nCount    <- rowSums(Y)
has_counts   <- df$nCount > 0
df           <- df[has_counts, ]
Y            <- Y[has_counts, , drop = FALSE]

# Technical FOV grouping for E^tech / within-section centering (NOT an RE block).
df$imageID <- factor(df$slide)

# Gene QC: keep genes detected in >= min_detection of cells of >= 1 cell type.
celltype_chr  <- as.character(df$celltype)
cells_by_type <- lapply(cell_types, function(ct) which(celltype_chr == ct))
max_detection <- vapply(seq_len(ncol(Y)), function(gene_i) {
  per_type <- vapply(cells_by_type, function(idx) {
    if (length(idx) == 0L) 0 else mean(Y[idx, gene_i] > 0)
  }, numeric(1))
  max(per_type)
}, numeric(1))

Y         <- Y[, max_detection >= min_detection, drop = FALSE]
df$nCount <- rowSums(Y)
cat(sprintf("%d cells x %d genes; %d slides\n",
            nrow(Y), ncol(Y), nlevels(df$imageID)))
```

# PACE fit (per-cell-HC, streaming pipeline)

The entire method — neighbour kernels (`K_bio` / `K_tech`), sparse-pair
dropping and within-section centering, the cross-celltype `E^tech`
ambient field with isotropic edge correction, the homotypic-core
negative-control anchors, data-informed τ, and the NB1 penalised
quasi-likelihood fit — is encapsulated in a single call to
**`pace_fit_streaming()`** (defined in `stage0/pace_core.R`). The
memory-bounded *streaming* solver never materialises the dense `E^tech`;
it streams `ambient_W %*% Y` instead.

`pace_fit_streaming()` is cohort-agnostic: BC, Mel, and any other dataset
call the **same** function and differ only in arguments. BC passes no
disease contrast and a single global section.

```{r pace, cache = TRUE}
res <- pace_fit_streaming(
  Y = Y, df = df, types = cell_types,
  celltype_col = "celltype", image_col = "slide", coord_cols = c("x", "y"),
  h_bio = h_bio, h_tech = h_tech, eps = kernel_radius,
  contamination     = "percell_hc",   # mu = mu_bio + rho_i * a_ig
  dispersion        = disp_model,      # canonical: NB1
  drop_sparse_neff  = min_neff,        # drop (focal, nb) cols with eff. n < 30
  within_image      = TRUE,            # centre kernels within (slide, celltype)
  edge_correct      = TRUE,
  data_informed_tau = TRUE,
  n_iter = n_iter, threads = pql_threads, tau_shrinkage = tau_shrinkage,
  verbose = TRUE
)

# Bind the pieces the reporting + figures read by name.
fit     <- res$fit
df      <- res$df            # working frame (kernel columns added)
Y       <- res$Y
X_fixed <- res$X_fixed
K_tech  <- res$K_tech
K_bio   <- res$K_bio
```

# mashr shrinkage + variance decomposition + MCSD

`pace_shrink()` runs mashr (Urbut 2019) over the neighbour-slope BLUPs.
`pace_decompose()` partitions within-focal latent variance into
Cell type / Spatial / Spillover / Residual (the 4-block view, plus the
raw 5-block table). `pace_top_drivers()` ranks genes per pair by
`b̃² · spec² · focal_mean` (lfsr < 0.05).

```{r downstream}
shrunken_long <- pace_shrink(fit, cell_types)
decomp        <- pace_decompose(fit, df, Y, cell_types, X_fixed)
gene_decomp   <- decomp$gene_focal_5block   # figures rank against the 5-block table
```

```{r mcsd}
# ALL ordered focal != neighbour pairs (omit `pairs=`). The four canonical
# focal<-Tumour pairs are only a reporting subset, not part of the method.
pair_drivers <- pace_top_drivers(fit, shrunken_long, decomp, cell_types)

for (pair_key in names(pair_drivers)) {
  top_genes <- head(pair_drivers[[pair_key]]$scores$gene, 8)
  cat(sprintf("%s  n_sig=%d  (%s): %s\n", pair_key,
              nrow(pair_drivers[[pair_key]]$scores),
              pair_drivers[[pair_key]]$status, paste(top_genes, collapse = ", ")))
}

# Bundle everything a downstream figure script would need.
mv <- list(
  fit = fit, shrunken_long = shrunken_long, gene_set = colnames(Y),
  decomposition = decomp, mcsd_canonical = pair_drivers,
  h_tech = h_tech, h_bio = h_bio, K_tech = K_tech, K_bio = K_bio,
  cell_meta = df[, c("cell_id", "nCount", "slide", "celltype")]
)
```

# Variance-component (τ) diagnostic

`τ` is the empirical-Bayes prior variance of each random-effect coefficient —
how much that *(focal × term)* effect varies across genes. The **`(Intercept)`
column is cell-type identity** (large τ); the **neighbour-slope columns are
mostly ≈0** (shrunk away), leaving a short tail of flagship spatial slopes. It's
a health-check: shrinkage is large where it should be (identity) and ≈0
elsewhere, so the slopes that survive are the credible spatial signals.

```{r tau-diagnostic, fig.width = 13, fig.height = 5}
p.tau <- plot_tau_diagnostic(fit)
```

<!-- The refit is persisted at the end of the notebook (see the "Save" section). -->

# Variance decomposition

**4-block VPC** — per focal cell type, the % of within-focal latent
variance in each block, pooled across genes (Goldstein 2002).

```{r block-vpc}
mv$decomposition$gene_focal_4block |>
  group_by(focal) |>
  summarise(across(
    c(`Cell type %`, `Spatial state %`, `Spillover %`, `Residual %`),
    ~ round(mean(.x), 1)
  )) |>
  arrange(desc(`Spatial state %`))
```

**Pratt pair attribution** — the Spatial block split into signed
`(focal, neighbour)` shares (suppressor neighbours allowed; shares sum
to the block total).

```{r pair-vpc}
pratt <- pace_pair_variance_pratt(
  mv, cond_prefix = NULL, cohort_label = "BC", block_label = "Spatial"
)
pratt$pair_long |>
  arrange(desc(within_focal_share_pct)) |>
  select(focal, neighbour, within_focal_share_pct) |>
  head(12)
```

# Top driver genes per pair

MCSD-ranked drivers (lfsr \< 0.05) for the four canonical BC pairs.
`b_clean` = mash-shrunken slope; `R2_S` = per-gene spatial variance
share.

```{r drivers}
for (pair_key in names(pair_drivers)) {
  cat("\n###", pair_key, "(", pair_drivers[[pair_key]]$status, ")\n")
  pair_drivers[[pair_key]]$scores |>
    head(10) |>
    select(rank, gene, MCSD, b_clean, spec, focal_mean, R2_S, lfsr) |>
    print()
}
```

# Decomposition figures

The panels from the canonical composite (`plot_bc313_composites_all.R`),
reproduced on this notebook's `mv`. **All panels use the canonical single-frame
decomposition** (observed CP10k denominator; the headline frame, spatial
1.8-5.8%, not the link-scale ~1% of `gene_focal_4block`). The stage-0
`pace_decompose()` does not build it, so we compute it here from the locked
4-block proportions with `single_frame_decomp_obs()`.

```{r single-frame}
# CANONICAL single-frame per-(gene,focal) decomposition (observed denominator).
mv$decomposition$gene_focal_single_frame <- single_frame_decomp_obs(
  Y, df$celltype, df$nCount, mv$decomposition$gene_focal_4block)
sf <- mv$decomposition$gene_focal_single_frame

# Pooled (over genes) single-frame block % per focal cell type.
focal_blocks <- sf |>
  mutate(ss_cell = `Cell type %` / 100 * denom,
         ss_sp   = `Spatial %`   / 100 * denom,
         ss_bl   = `Spillover %` / 100 * denom,
         ss_rs   = `Residual %`  / 100 * denom) |>
  group_by(focal) |>
  summarise(`Cell type`          = 100 * sum(ss_cell) / sum(denom),
            `Spatial cell state` = 100 * sum(ss_sp)   / sum(denom),
            `Spillover`          = 100 * sum(ss_bl)   / sum(denom),
            `Residual`           = 100 * sum(ss_rs)   / sum(denom),
            .groups = "drop")
```

**Per-focal decomposition (bar + zoom)** via the locked
`plot_stacked_bar_focal_no_resp()` (full stack + a zoom of the small Spatial +
Spillover blocks), single-frame values.

```{r decomp-bar, fig.width = 13, fig.height = 5}
blks_all <- focal_blocks |>
  transmute(focal,
            `Cell type`          = `Cell type` / 100,
            `Spatial cell state` = `Spatial cell state` / 100,
            `Spillover`          = `Spillover` / 100,
            `Residuals`          = `Residual` / 100) |>
  pivot_longer(-focal, names_to = "block", values_to = "pct_total")

p.decomp <- plot_stacked_bar_focal_no_resp(
  blks_all,
  block_levels = c("Cell type", "Spatial cell state", "Spillover", "Residuals"),
  title = "Per-focal decomposition")
```

**Pairwise spatial % heatmap.** Each cell = the % of the focal type's TOTAL
variance contributed by spatial interaction with that neighbour: the focal's
single-frame spatial % (panel above) split across neighbours by the
off-diagonal-renormalised Pratt share (`foc_sp × V_pair / Σ V_pair`). Matches
`plot_bc313_composites_all.R` panel C.

```{r decomp-heatmap, fig.width = 6, fig.height = 4}
TYPES  <- sort(unique(as.character(df$celltype)))
foc_sp <- setNames(focal_blocks$`Spatial cell state`,
                   as.character(focal_blocks$focal))

pratt_long <- pace_pair_variance_pratt(
  mv, cond_prefix = NULL, cohort_label = "BC", block_label = "Spatial")$pair_long |>
  filter(as.character(focal) != as.character(neighbour)) |>
  group_by(focal) |>
  mutate(val = foc_sp[as.character(focal)] * V_pair_pratt / sum(V_pair_pratt)) |>
  ungroup() |>
  transmute(focal = as.character(focal),
            neighbour = as.character(neighbour), val)

# Fill the full focal x neighbour grid (undefined/dropped pairs -> 0; diagonal blank).
heat <- expand.grid(focal = TYPES, neighbour = TYPES, stringsAsFactors = FALSE) |>
  filter(focal != neighbour) |>
  left_join(pratt_long, by = c("focal", "neighbour")) |>
  mutate(val       = ifelse(is.na(val), 0, val),
         focal     = factor(focal, levels = TYPES),
         neighbour = factor(neighbour, levels = rev(TYPES)))

p.heat <- ggplot(heat, aes(focal, neighbour, fill = val)) +
  geom_tile(colour = "grey92", linewidth = 0.3) +
  geom_text(aes(label = ifelse(abs(val) < 0.005, "0.00", sprintf("%.2f", val))),
            size = 2.4, colour = "white") +
  scale_fill_viridis_c(option = "inferno", na.value = "white",
                       name = "spatial %\nof total") +
  coord_equal() +
  labs(x = "Focal", y = "Neighbour",
       title = "Pairwise spatial % (variance, normalised Pratt)") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
```

**Strong/weak split of the Pratt pair variance.** The pairwise Pratt above pools
a covariance sum over all genes, so it is far less null-diluted than a mean of
per-gene ratios. To make that explicit (and null-robust by construction) we split
each pair's per-gene Pratt contribution into a STRONG component carried by the
significant driver genes (lfsr < 0.05) and a WEAK component carried by the rest,
following the strong-localized / weak-diffuse partition of Nan-Azriel-Schwartzman
(2026). STRONG + WEAK reproduces the total above; the STRONG panel is the
null-robust fill and WEAK is the diagnostic of how much of each pair was
null-carried. `pace_pair_variance_pratt_strong_weak()` is auto-sourced from
`scripts/helpers/`; it does not modify the locked Pratt helper.

```{r decomp-heatmap-strong-weak, fig.width = 6, fig.height = 4}
# Same foc_sp scaling as the total-Pratt heatmap above, applied to each component.
pratt_sw <- pace_pair_variance_pratt_strong_weak(
  mv, cond_prefix = NULL, cohort_label = "BC") |>
  filter(as.character(focal) != as.character(neighbour)) |>
  group_by(focal) |>
  mutate(denom      = sum(V_pair_pratt),
         strong_pct = foc_sp[as.character(focal)] * V_pair_strong / denom,
         weak_pct   = foc_sp[as.character(focal)] * V_pair_weak   / denom,
         pair_pct   = foc_sp[as.character(focal)] * V_pair_pratt  / denom) |>
  ungroup() |>
  transmute(focal = as.character(focal), neighbour = as.character(neighbour),
            n_strong, pair_pct, strong_pct, weak_pct,
            pct_from_strong = 100 * V_pair_strong / V_pair_pratt)

# Table: top pairs by the null-robust STRONG share.
pratt_sw |>
  arrange(desc(strong_pct)) |>
  mutate(across(c(pair_pct, strong_pct, weak_pct, pct_from_strong), \(x) round(x, 3))) |>
  head(12) |>
  knitr::kable(caption = "BC Pratt pair % split: STRONG (drivers) vs WEAK (nulls)")

# STRONG-filled heatmap (null-robust); mirrors the total-Pratt heatmap style.
heat_strong <- expand.grid(focal = TYPES, neighbour = TYPES, stringsAsFactors = FALSE) |>
  filter(focal != neighbour) |>
  left_join(pratt_sw |> select(focal, neighbour, strong_pct), by = c("focal", "neighbour")) |>
  mutate(strong_pct = ifelse(is.na(strong_pct), 0, strong_pct),
         focal      = factor(focal, levels = TYPES),
         neighbour  = factor(neighbour, levels = rev(TYPES)))

p.heat_strong <- ggplot(heat_strong, aes(focal, neighbour, fill = strong_pct)) +
  geom_tile(colour = "grey92", linewidth = 0.3) +
  geom_text(aes(label = ifelse(abs(strong_pct) < 0.005, "0.00", sprintf("%.2f", strong_pct))),
            size = 2.4, colour = "white") +
  scale_fill_viridis_c(option = "inferno", na.value = "white",
                       name = "spatial %\nof total\n(drivers)") +
  coord_equal() +
  labs(x = "Focal", y = "Neighbour",
       title = "Pairwise spatial % — strong-localized (driver-gene) Pratt") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
p.heat_strong

# Persist the strong/weak decomposition on the fit object (saved at the end).
mv$pratt_strong_weak <- pratt_sw
```

**Ligand-receptor annotation of the top pairs (CellChatDB).** An interpretation
layer on the drivers: for each top focal<-neighbour pair, every significant
driver is checked against CellChatDB; if it is a ligand or a receptor subunit,
the cognate partner is named and its expression in the NEIGHBOUR cell type is
reported, turning the spatial coupling into a candidate signaling axis. The
313-panel measures only a handful of complete LR pairs, so this is a targeted
annotation, not an LR screen. Because the drivers are contamination-corrected,
an axis built on them is cleaner than raw spatial LR co-expression (which is
confounded by segmentation spillover). `pace_lr_annotate()` is auto-sourced from
`scripts/helpers/`; homotypic/adhesion pairs (CDH1, PECAM1) are flagged, and
one-sided hits (cognate partner off-panel) are counted but not tabulated. The
table is ordered by `driver_pctile` — how prominent the driver is among the
pair's significant drivers by |slope| — so the axes that genuinely reflect the
PACE result (e.g. EDN1 in Endothelial, rank 1; CXCL12/MRC1 in Macrophage, ranks
3/5) sit at the top and marginal tail hits are visible as such. `direction`
states whether the gene rises or falls with neighbour density; several strong
axes are *down*-near-tumour, so they should be read as *suppression* in
proximity, not an activated signaling axis.

```{r lr-annotation}
if (requireNamespace("CellChat", quietly = TRUE)) {
  top_pairs_bc <- pratt_sw |> dplyr::arrange(dplyr::desc(strong_pct)) |> head(6)
  pair_list    <- Map(function(f, n) c(f, n), top_pairs_bc$focal, top_pairs_bc$neighbour)
  lr_bc        <- pace_lr_annotate(mv, pair_list)
  mv$lr_axes   <- lr_bc                                  # persisted with the saved fit
  adhesion     <- unique(lr_bc$driver[lr_bc$class != "paracrine"])
  cat(sprintf("Adhesion/homotypic (flagged): %s  |  one-sided candidates (partner off-panel): %d\n",
              paste(adhesion, collapse = ", "), sum(!lr_bc$partner_on_panel)))
} else {
  message("CellChat not installed - skipping ligand-receptor annotation."); lr_bc <- NULL
}

if (!is.null(lr_bc)) {
  lr_bc |>
    dplyr::filter(partner_on_panel, class == "paracrine") |>
    dplyr::arrange(dplyr::desc(driver_pctile)) |>          # strongest-PACE-driver axes first
    dplyr::transmute(pair = paste0(focal, " <- ", neighbour), driver, role, direction,
                     driver_rank, partner, category, partner_where, partner_spec) |>
    knitr::kable(caption = "Candidate LR axes behind BC top pairs, ranked by PACE-driver prominence (CellChatDB). category = CellChatDB interaction type; partner_where = compartment expressing the cognate partner (autocrine = focal, paracrine = neighbour); partner_spec = partner expression there relative to its cross-cell-type max. The dot plot below keeps only Secreted-Signaling axes with partner_spec >= 0.2.")
}
```

Standard ligand-receptor **dot plot** (CellPhoneDB / LIANA style): focal<-neighbour
pairs on x, ligand->receptor axes on y, dot **size = PACE driver prominence**
(percentile) and **colour = driver slope** (blue = down / red = up near the
neighbour). Bidirectional axes are collapsed to the more prominent side. Two
credibility filters keep it to plausible signaling (both adjustable in
`pace_lr_dotplot()`): only **Secreted-Signaling** CellChatDB interactions
(dropping the soft Cell-Cell-Contact set, e.g. the spurious PTPRC->MRC1 where
MRC1 is a macrophage receptor absent from T cells), and only axes whose cognate
partner is genuinely expressed in the focal or neighbour (`partner_spec >= 0.2`,
direction-aware so real autocrine axes such as EDN1->EDNRB survive).

```{r lr-dotplot, fig.width = 8.5, fig.height = 5.5}
if (!is.null(lr_bc)) {
  print(pace_lr_dotplot(lr_bc, "Candidate ligand-receptor axes behind BC top spatial couplings"))
}
```

**MCSD drivers + per-gene variance for one pair.** Pick a focal-neighbour pair
(default the Stromal ← Tumour flagship): the left bar ranks its top genes by
MCSD, the right shows each gene's single-frame block decomposition (% of total
variance; residual is the implied remainder).

```{r decomp-pair-genes, fig.width = 12, fig.height = 5}
# make_pair_composite() lives in scripts/core/06a-plots-decomp.R (sourced in setup).
# panels = "both" | "mcsd" | "gene"; change the pair to render any mv$mcsd_canonical pair.
p_stro_tum <- make_pair_composite("Stromal", "Tumour", mv, sf, n_top = 5)
p_mac_tum <- make_pair_composite("Macrophage", "Tumour", mv, sf, n_top = 5)
p_bcell_tcell <- make_pair_composite("B_Cell", "T_Cell", mv, sf, n_top = 5)
p_tum_myoepi <- make_pair_composite("Tumour", "Myoepithelial", mv, sf, n_top = 5)
```

# Contamination diagnostics

The per-cell-HC model estimates one loading `ρ_i` per cell (shared
across genes, fixed ambient direction). Its distribution + the Spillover
variance share show how much technical bleed the model absorbed, keeping
it out of the spatial slopes.

```{r contam}
rho <- fit$percell_bleed_rho
cat(sprintf("rho_i: median %.4f, IQR [%.4f, %.4f], frac>0 %.2f\n",
            median(rho), quantile(rho, .25), quantile(rho, .75), mean(rho > 0)))

p_rho <- ggplot(data.frame(rho = rho), aes(rho)) +
  geom_histogram(bins = 60, fill = "#4292C6", colour = NA) +
  geom_vline(xintercept = median(rho), linetype = 2) +
  labs(x = expression(rho[i] ~ "(per-cell contamination loading)"),
       y = "cells", title = "Per-cell contamination loading") +
  theme_bw()

spillover_by_focal <- mv$decomposition$gene_focal_4block |>
  group_by(focal) |>
  summarise(`Spillover %` = mean(`Spillover %`))

p_spill <- ggplot(spillover_by_focal,
                  aes(reorder(focal, `Spillover %`), `Spillover %`)) +
  geom_col(fill = "#E6550D") +
  coord_flip() +
  labs(x = NULL, y = "Spillover variance share (%)",
       title = "Spillover block by focal type") +
  theme_bw()

p_rho | p_spill
```

# MCSD neighbour-count boxplots

For the top two pairs, the top MCSD gene's raw counts across bins of the
number of tumour neighbours within 30 µm of each focal cell — the spatial
gradient the slope captures. Box = distribution, dot = per-bin mean.

```{r boxplots, fig.width = 10, fig.height = 4}
# density_bin_boxplot() / top_gene() are in helpers/notebook_analysis_helpers.R

p_den_stro_tum <- density_bin_boxplot("Stromal", "Tumour", gene_name = c("ADH1B", "CCDC80", "CXCL12", "SFRP4"))
p_den_mac_tum <- density_bin_boxplot("Macrophage", "Tumour", gene_name = c("MRC1", "CD14", "APOC1", "C1QA"))
p_den_bcell_tcell <- density_bin_boxplot("B_Cell", "T_Cell", gene_name = c("TENT5C", "MZB1", "MS4A1", "PRDM1"))
p_den_tum_myoepi <- density_bin_boxplot("Tumour", "Myoepithelial", gene_name = c("ABCC11", "SERHL2", "FASN", "FOXA1"))
```

# Spatial inset figures

Cell-type spatial map with two tumour-duct ROIs zoomed as insets, focal
cells heat-coloured by gene expression. Uses the canonical
`compose_inset_figure` helper. Shows the top two genes for the top three
pairs.

```{r insets-data}
# compose_inset_figure() is auto-sourced in the setup chunk.

# Plotting df: coords + cell type + log-CP10K for the genes we inset.
plot_df <- data.frame(x = df$x, y = df$y, celltype = as.character(df$celltype))
inset_genes <- c("ADH1B", "CCDC80", "MRC1", "APOC1", "ABCC11", "SERHL2")
inset_genes <- inset_genes[inset_genes %in% colnames(Y)]
for (gene in inset_genes) {
  plot_df[[gene]] <- log1p(1e4 * Y[, gene] / df$nCount)
}

# Two right-side tumour-ring ROIs (box 1 = orange, box 2 = blue).
roi1_x <- c(6300, 7300); roi1_y <- c(3700, 4700)
roi2_x <- c(6500, 7500); roi2_y <- c(2300, 3300)
```

```{r inset-stromal, fig.width = 13, fig.height = 7}
if (all(c("ADH1B", "CCDC80") %in% colnames(plot_df))) {
  p_inset_stro_tum <- compose_inset_figure(
    plot_df, genes = c("ADH1B", "CCDC80"),
    focal = "Stromal", neighbour = "Tumour",
    roi1_x, roi1_y, roi2_x, roi2_y
    # roi_col1 = "darkorange", roi_col2 = "steelblue"
  )
}
```

```{r inset-mac, fig.width = 13, fig.height = 7}
if (all(c("MRC1", "APOC1") %in% colnames(plot_df))) {
  p_inset_mac_tum <- compose_inset_figure(
    plot_df, genes = c("MRC1", "APOC1"),
    focal = "Macrophage", neighbour = "Tumour",
    roi1_x, roi1_y, roi2_x, roi2_y
    # roi_col1 = "darkorange", roi_col2 = "steelblue"
  )
}
```

```{r inset-tum, fig.width = 13, fig.height = 7}
if (all(c("ABCC11", "SERHL2") %in% colnames(plot_df))) {
  p_inset_tum_myoepi <- compose_inset_figure(
    plot_df, genes = c("ABCC11", "SERHL2"),
    focal = "Tumour", neighbour = "Myoepithelial",
    roi1_x, roi1_y, roi2_x, roi2_y
    # roi_col1 = "darkorange", roi_col2 = "steelblue"
  )
}
```

# Save the PDFs
```{r}
pdf("../figures/bc/all_plots.pdf", width = 16, height = 8)
p.tau
p.decomp
p.heat
p_stro_tum
p_mac_tum
p_bcell_tcell
p_tum_myoepi
p_den_stro_tum
p_den_mac_tum
p_den_bcell_tcell
p_den_tum_myoepi
p_inset_stro_tum
p_inset_mac_tum
p_inset_tum_myoepi
dev.off()
```


# Save

Persist this notebook's refit — including the strong/weak Pratt decomposition
(`mv$pratt_strong_weak`) attached above — to `output_path`. This writes a
notebook-specific file (`..._notebook.rds`), NOT the canonical fit.

```{r save-fit}
saveRDS(mv, "../data/breast_cancer/canonical_fit.rds")
```

# Session info

```{r session}
sessionInfo()
```
