---
title: "PACE melanoma analysis (Figure 4)"
author: "Elijah Willie"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

This notebook reproduces the melanoma analysis and Figure 4 of the PACE
manuscript **using the [PACE package](https://github.com/ecool50/PACE)**. Melanoma
is a multi-sample, condition-stratified cohort (progressive vs non-progressive
disease after immune checkpoint blockade), so the fit adds a responder spatial
block; the package handles this through `condition_col` / `resp_term`. Rendered
from the repository root:

```bash
quarto render pace_mv/notebooks/analysis_mel.qmd
```

# Setup

```{r setup}
library(PACE)
library(SpatialExperiment)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(zellkonverter)
library(dplyr)
library(forcats)
library(ggplot2)
library(Matrix)

# canonical melanoma knobs
h_bio <- 30; h_tech <- 5; px_size <- 0.12028; min_detection <- 0.05
cell_types      <- c("Tumour", "Endothelial", "Fibroblast", "Macrophage",
                     "B_Cell", "T_Cell")
tumour_subtypes <- c("Tumor_a", "Tumor_b", "Tumor_c", "Tumor_d", "Tumor_e", "Tumor_f")
resp_ref <- "nonPD"; resp_case <- "PD"; resp_term <- paste0("Responder", resp_case)

h5ad_path <- "../../data/simvi_melanoma/14708000/Melanoma_5612.h5ad"   # fetched by fetch_data.R
```

# Load and prepare the cohort

Read the CosMx `.h5ad`, keep all recorded responses (progressive = PD; non-progressive
= SD + PR + CR), collapse to six cell types, keep genes detected in at least 5% of
some cell type, and set the responder factor with non-PD as the reference. This
reproduces the manuscript cohort (56,274 cells x 927 genes across 26 patients).

```{r prepare}
sce <- readH5AD(h5ad_path, reader = "R")
assayNames(sce)[assayNames(sce) == "X"] <- "counts"
keep <- sce$BEST_RESPONSE_BY_SCAN %in% c("PD", "SD", "CR", "PR")
sce  <- sce[, which(keep)]
sce$BEST_RESPONSE_BY_SCAN <- ifelse(
  as.character(sce$BEST_RESPONSE_BY_SCAN) == "PD", "PD", "nonPD")

df <- cbind(colData(sce), t(as.matrix(assay(sce, "counts"))),
            reducedDim(sce, "spatial")) |>
  as.data.frame() |>
  mutate(imageID = paste(SPID, fov, sep = "_"),
         x = V1 * px_size, y = V2 * px_size,
         celltype = make.names(celltype),
         cell_ID = paste0(cell_ID, "_", fov))
df$celltype <- recode(df$celltype, `B.cell` = "B_Cell", `T.CD8.memory` = "T_Cell",
  `T.CD8.naive` = "T_Cell", `Treg` = "T_Cell", `endothelial` = "Endothelial",
  `fibroblast` = "Fibroblast", `macrophage` = "Macrophage")
df$celltype[grepl("^T\\.", df$celltype)] <- "T_Cell"
df$celltype <- ifelse(df$celltype %in% tumour_subtypes, "Tumour", df$celltype)
df <- df[df$celltype %in% cell_types, ]
df <- df[order(df$imageID), ]                       # per-image kernel / RE row order
panel_genes <- make.names(rownames(sce))

# gene QC (detection >= 5% in some cell type), drop zero-count cells
genes <- intersect(panel_genes, names(df)); ctv <- as.character(df$celltype)
cbt <- lapply(cell_types, function(t) which(ctv == t))
det <- vapply(genes, function(g)
  max(vapply(cbt, function(ix) mean(df[ix, g] > 0), numeric(1))), numeric(1))
genes <- genes[det >= min_detection]
df$nCount <- rowSums(as.matrix(df[, genes, drop = FALSE]))
df <- df[df$nCount > 0, ]

df$celltype  <- factor(as.character(df$celltype), levels = cell_types)
df$imageID   <- factor(as.character(df$imageID))
df$Responder <- fct_relevel(as.factor(df$BEST_RESPONSE_BY_SCAN), resp_ref)  # nonPD ref

# assemble the SpatialExperiment
Y <- t(as.matrix(df[, genes, drop = FALSE])); rownames(Y) <- genes
cn <- make.unique(as.character(df$cellID_str)); colnames(Y) <- cn
spe <- SpatialExperiment(
  assays = list(counts = Y),
  colData = DataFrame(cellType = df$celltype, Responder = df$Responder,
                      imageID = df$imageID, row.names = cn),
  spatialCoords = as.matrix(df[, c("x", "y")]))

sprintf("%d cells x %d genes across %d patients", ncol(spe), nrow(spe),
        nlevels(spe$imageID))
table(spe$Responder)
```

# Fit PACE (responder model)

`paceFit()` runs the same pipeline as breast cancer, with the disease contrast
(`condition_col`), per-patient kernels (`kernel_per_image = TRUE`), and an image
random intercept. `resp_term = "ResponderPD"` flags PD as the case (non-PD is the
reference).

```{r fit, cache = TRUE}
fit <- paceFit(spe,
               celltype_col      = "cellType",
               image_col         = "imageID",
               types             = cell_types,     # canonical block order
               condition_col     = "Responder",
               resp_term         = "ResponderPD",
               contamination     = "percell_hc",
               dispersion        = "nb1",
               h_bio = 30, h_tech = 5,
               kernel_per_image  = TRUE,
               image_re          = "intercept",
               drop_sparse_neff  = 30,
               within_image      = TRUE,
               edge_correct      = TRUE,
               data_informed_tau = TRUE,
               n_iter = 32L, threads = 4L, tau_shrinkage = "adaptive",
               verbose = FALSE)
fit
```

SPP1 is the only gene reaching significance for the responder component of the
macrophage–tumour interaction (manuscript slope ≈ −0.067):

```{r spp1}
neighbourSlopes(fit) |>
  filter(focal == "Macrophage", neighbour == "Tumour",
         term == "ResponderPD:Tumour", lfsr < 0.05) |>
  select(gene, estimate_shrunk, lfsr)
```

# Figure 4a — per-focal decomposition (with responder block)

```{r fig4a, fig.width = 13, fig.height = 5.5}
plotDecomposition(fit, title = "Melanoma")
```

Macrophages and fibroblasts carry the largest spatial contributions; the responder
spatial state is small across all populations, and contamination is highest in the
sparsely distributed immune and stromal populations.

# Figure 4b — responder-by-proximity heatmap

```{r fig4b, fig.width = 6.5, fig.height = 5}
plotPairHeatmap(fit, block = "responder")
```

# Figure 4c — macrophage SPP1 versus tumour proximity, by outcome

The manuscript panel splits macrophage SPP1 by response. Here we bin each
macrophage by its number of tumour neighbours and show the mean SPP1 (raw counts)
per bin, separately for progressive and non-progressive disease. This is a
descriptive view: SPP1 rises steeply with tumour proximity in both outcomes, and
progressive-disease macrophages are concentrated at high tumour proximity with the
highest SPP1 — the population-level SPP1+ macrophage signal PACE flags. The
model-based responder slope (adjusting for baseline state and contamination, above)
is the formal estimate; the raw means here still confound baseline cell-state level,
which is where most of the response difference sits.

```{r fig4c, fig.width = 6.5, fig.height = 4}
coords <- spatialCoords(spe); ct <- spe$cellType
mac <- which(ct == "Macrophage")
nn  <- dbscan::frNN(coords[which(ct == "Tumour"), , drop = FALSE], eps = 30,
                    query = coords[mac, , drop = FALSE])
d <- data.frame(
  tum_nbr  = cut(lengths(nn$id), c(0, 2, 4, 6, 8, Inf), include.lowest = TRUE),
  SPP1     = as.numeric(assay(spe, "counts")["SPP1", mac]),
  Response = ifelse(spe$Responder[mac] == "PD", "progressive", "non-progressive"))

ggplot(d, aes(tum_nbr, SPP1, colour = Response, group = Response)) +
  stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
  stat_summary(fun = mean, geom = "point", size = 2.5) +
  scale_colour_manual(values = c(progressive = "#D62728", `non-progressive` = "#1F77B4")) +
  labs(x = "Number of tumour neighbours", y = "Macrophage SPP1 (raw counts, mean)",
       title = "SPP1 vs tumour proximity by outcome") +
  theme_minimal(base_size = 12)
```

The per-pair driver composite (macrophage–tumour) and the raw proximity boxplot:

```{r fig4c-drivers, fig.width = 10, fig.height = 4.5}
plotDrivers(fit, "Macrophage", "Tumour")
plotProximity(fit, spe, "SPP1", "Macrophage", "Tumour")
```

# Supplementary — contamination anchor genes

```{r anchors}
anchorGenes(fit, spe)
```

# Session info

```{r sessioninfo}
sessionInfo()
```
